---
title: "STAT 545 B Assignment B1"
author: "Gopal Khanal"
date: "2023-11-03"
output: github_document
---

# Introduction

This is the first assignment of UBC's STAT 545 course Part B. This assignment covers making a function in R, documenting it, and testing it. This includes four specific exercises. 
  1. Making a reasonable function.
  2. Documenting the function.
  3. Showing the application of the function with example
  4. Demonstrating if the function actually works with tests.

# Excercise 1 & 2 : Create a function and its documentation


First, I will load the relevant libraries that I'll need.

```{r, message = FALSE}
library(datateachr) # <- This  contains that the data flow sample I'm using
library(palmerpenguins) # <- This  contains that the data flow sample I'm using
library(tidyverse)
library(testthat)
library(roxygen2)
library(dplyr)
```

One of the main tasks in data analysis is that we often need to check how observations of a particular variable vary across other variables. For example, how blood pressure or hemoglobin content of patients vary across different treatment groups, which could be age, sex, or other treatment levels.

While this can be easily done in base R or using `dplyr` package, creating a `function` could ease the task when one wants to do the things repeatedly.

In my previous Mini Data Analysis, I had to regularly check and repeat how **flow** (continuous variable) varied across extreme **flow type** (categorical variable). This is often the case in other types of analysis as well when one need to constantly check how observation values vary across categories or treatment types. 

My function `summary.stat`computes summary statistics for a numeric variable within a data frame, grouped by another variable. The grouping variable can be categorical, character variable or numeric. Ideally, it's better to do such summary for a grouping  variable with categorical or character class. But, users may also want to summarize one numeric variable across other variable with numeric or integer class. This function summarizes the value different summary statistics such as mean, range, standard deviation,  of the continuous variable across other variables.

```{r}

# Naming the function as summary.stat
summary.stat <- function(dataset, group, value, na.rm = TRUE) {
  # Make the column names as string so that they are  used checking error and getting for error messages later.
    group1 <- deparse(substitute(group))
  value1 <- deparse(substitute(value))
    #  Check if there are non-numeric values in the variable that we want to summarize
   if(!is.numeric(dataset[[value1]])) {
    stop('Sorry, numeric variable can only be a numeric vector')
   }
  # check if there are  NAs in the grouping variable. If yes, NAs will be considered and summarized as separate group.
    if((!is.null(group1) && anyNA(dataset[[group1]]))){
    warning("NAs are found in grouping variable. Consider NAs as a separate group")
  }
  # Implement the function
  dataset %>%
    group_by({{group}}) %>%
    summarize(
      n = n(), 
      mean = mean({{value}},na.rm = T),
      range= max({{value}},na.rm = T)-min({{value}},na.rm = T),
      stdev = sd({{value}},na.rm = T))
}
#'@Title Get summary statistics of a variable across other variable
#' @Description: This function returns a table with the summary statistics such as total number of observations, mean, range and standard deviation of continuous variable across other variable.
#' @param dataset A data frame that should have at least one continuous variable and other grouping variables. This is named to make better sense.
#' @param group A vector which could be of factor, character or numeric or integer. The function will group values of another numeric variable according to this variable to yield summary statistics with new function. This is named as group to make typical sense about the grouping variable.
#' @param value A numeric vector which should have numeric class. I named it value because it will be self evident that this variable is the one for which we have observation/values to summarize. 
#' @param na.rm Logical, indicating whether to remove missing values (default is TRUE).
#'
#' @return A data frame with summary statistics information given by this new function.
#' @export
#'
#' @examples
#' (summary.stat(flow_sample, flow, extreme_type))
#' 
```

# Exercise 3: Showing how function works with examples 

## Flow sample data

The **flow_sample** data that I used for Mini Data Analysis had **flow** as continuous variable and **extreme_type** as categorical variable two two level. Below I show application of the new function **summary_stat** to quickly access the summary information of different summary measures (number of observations, mean, standard deviation and other)

```{r}

flow_summary <- summary.stat(flow_sample, extreme_type, flow)

flow_summary

```


## Vancouver tree data

The **Vancouver tree** data has tree information for different neighborhood. One important measure of tree is diameter, and the tree **diameter** can vary based on categorical variable **neighborhood_name**. The **summary.stat** function can quickly summarise selected tree summary measures for different neighborhoods.

```{r}
tree_summary <- summary.stat(vancouver_trees, neighbourhood_name, diameter)
tree_summary
```


## Penguins data set

The **penguins** data has body mass information of different species of penguins. It is important to get summary information of **body_mass_g** for different **species** (categorical variable) of penguins. 
```{r}
penguins_summary <- summary.stat(penguins, species, body_mass_g)
penguins_summary
```

## Mtcars data
### Use of numeric class variable as grouping variable

Not-specifying grouping variable as character or factor works for numeric or integer class variable. For example, here I have summarized **drat** across **cyl** which has numeric class.
```{r}
class(mtcars$cyl)
mpg_summary <- summary.stat(mtcars, cyl, drat)
mpg_summary

```


# Exercise 4: Testing the function

Write formal tests for your function. You should use at least three non-redundant uses of an expect_() function from the testthat package, and they should be contained in a test_that() function (or more than one). They should all pass.

The **function** should work in range of conditions.  Below I use package `testthat` to test if this function actually works in different set of conditions.


```{r}
library(testthat)

```

### Test 1: Check whether the function output matches with the normal calculation

```{r}
# Use dplyr package to get the summary output of the data and rename the dataframe

data1 <- penguins %>%
  group_by(species) %>%  
  summarize(n=n(),
    mean = mean(body_mass_g, na.rm = TRUE),
    range=max(body_mass_g, na.rm = TRUE)- min(body_mass_g, na.rm=TRUE),
    stdev = sd(body_mass_g, na.rm = TRUE))
  
 # Check whether the output from function matches with the output from dplyr
test_that("Check if function gives correct calculation", {
  expect_equal(summary.stat(penguins, species, body_mass_g), data1) 
 
})
```

### Test 2: Test of incorrect class of input variables in generated data

```{r}

# Create sample data
value <- rnorm(100, mean = 0, sd = 1)
group <- sample(c('A', 'B'), 100, replace = TRUE)

# Create a dataframe with character and numeric variables
df <- data.frame(value = value, group = group)

# Display the dataframe
#print(df)

# check error with character class in place of numeric class

#expect_error(summary_stat(df, value, group))

test_that('Test other cases that should give an error', {
  expect_error(summary.stat(df, value, group))  
})


```
  

### Test 3: Use of incorrect class of input variables.

I'm using penguins dataset to get summary information. 

The function will  work in a dataset where my grouping variable is of any class factor, character or numeric, and the summarizing variable should be of numeric class.

My function `summary.stat` throws an error if my summarizing variable is of class other than numeric.

In the **vancouver_tree** data street address is non-numeric so it gives an error.
  
```{r}

result1<-'Sorry, numeric variable can only be a numeric vector'
test_that('Sorry, numeric variable can only be a numeric vector', {
expect_error(summary.stat(vancouver_trees, neighbourhood_name, std_street), result1)
})

```

### Test 4: Incorrect naming of input variables.

I used the incorrect name of the summarizing variable from other data set.

The function should through an error for this.

```{r}
# Test of incorrect variable name when variable from other data is used 
test_that('Test of incorrect variable name should give an error', {
  expect_error(summary.stat(flow_sample, 'diameter'))  
})
```
### Test 5: Test when the data contains NA's. The function will work in that case if NAs are handled by function.

For example, some of the values in penguins dataset has NAs. The function will handle the NA's and won't through any error. I'll calculate summary ignoring NAs.

```{r}

# Create a sample dataframe with NAs
test_data <- data.frame(
  species = c("A", "B", "A", "B", "A", "C", "C", "C"),
  body_mass_g = c(50, 60, NA, 70, 80, 90, NA, 30)
)
result2 <- test_data %>%
  group_by(species) %>%  
  summarize(n=n(),
    mean = mean(body_mass_g, na.rm = TRUE),
    range=max(body_mass_g, na.rm = TRUE)- min(body_mass_g, na.rm=TRUE),
    stdev = sd(body_mass_g, na.rm = TRUE))

test_that("Check if function gives correct calculation when the data contains NA's",{
expect_equal(summary.stat(test_data, species, body_mass_g), result2) })
  
```
